A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)

### 2018-04-12 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
updates

\title{A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)}
\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes} 
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

\red{Motivation for the work? What is the main aim of the work?} In this paper, a method for applying SVD to reduced the order of (\ie upscale) the permeability field (of a BaseCase) was introduced. This method was compared to arithmetic and harmonic averaging (\ie upscaling) technique as well as a stochastic probability density function (PDF) method of upscaling, to highlight its benefits. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase. Additionally, the SVD upscaling technique did not require a priori understanding of similar permeability blocks within the existing domain (\ie the BaseCase).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that rock properties such as permeability, pore spaces ({\it i.e.}, porosity) and pore throat vary spatially at different length scales. The heterogeneity of rock formations (\ie porous media) is relevant in calculations of transport and storage of fluids in porous media (\ie Multiphase flow in subsurface rocks). Therefore, the representation of these heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

In essence, heterogeneity makes it impossible to obtain reliable and accurate spatial information about geological properties throughout the domain ({\it i.e.}, uncertainty problems). Additionally, due to limitations in computing resources, it is neither efficient nor possible to use exact values of geological properties at all spatial coordinates of the domain for running fine-grid simulations (\ie scale problems) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. It is possible to overcome these challenges by upscaling. Upscaling techniques solve both problems by replacing discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve the statistical properties and flow dynamics behaviour of the high resolution domain give the best representation of the heterogeneous property. There are several upscaling techniques which fall under different categories \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}. Traditionally, the techniques were deterministic, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic upscaling techniques are robust enough to handle multiphase flow in porous media, and they address scale and uncertainity problems which were previously mentioned.

Upscaling is a term used in a few industries, some of which were mentioned in the first paragraph. It is more popularly known in the larger science and engineering community as a form of Model Order Reduction (MOR), or the output (\ie Reduced Order Model, ROM). Depending on the context in which MOR is being used, it can simply be referred to as dimensionality reduction. The definition of MOR also depends on the context it is being used \cite{Schilders2008}, but it suffices to state that MOR captures the essential features of a system, structure or parameter. This implies that an original system, structure or parameter exists with more detailed information. Fig.~\ref{fig:IllustrationMOR} presents a graphical intepretation of the MOR concept. It shows that even with the reduced number of facets, the graphic on the right of this illustration is recognized as that of a rabbit.

>>>>

\title{A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)}
\author[UoA]{B. Lashore}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes\corref{cor1}}\ead{jefferson.gomes@abdn.ac.uk}
%\author[UoA]{B. Lashore\corref{cor1}}\ead{lashorebabatunde@yahoo.com} \author[UoA]{K. Christou} \author[UoA]{J.L.M.A. Gomes} 
\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Research Group \\ School of Engineering, University of Aberdeen, UK}

@@

\red{Motivation for the work? What is the main aim of the work?} In this paper, a SVD-based method was developed to reduce the order of porous media permeability field (\ie upscale). This method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits \red{(performance??)}. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the SVD upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Introduction}\label{section:intro}
Naturally occuring rock formations are inherently heterogeneous, which means that rock properties such as permeability, pore spaces (\ie porosity) and pore throat vary spatially at different length scales. Heterogeneity of rock formations is relevant in calculations of transport and storage of fluids in porous media (\ie multiphase flow in subsurface rocks). Therefore, the representation of such heterogeneous properties is of great importance to oil and gas (\ie reservoir management), carbon capture and storage (CCS, \ie CO$_2$ transportation and storage in subsurface rocks) and waste management (\ie remediation of contaminated soil).

In essence, rock heterogeneity makes it impossible to obtain reliable and accurate spatial information about geological properties throughout the domain (\ie uncertainty problem). Additionally, due to limitations in computing resources, it is neither efficient nor possible to use exact values of geological properties at all spatial coordinates of the domain to perform fine-grid simulations (\ie scaling problem) \cite{chen_2006,miller_1998,Renard_1997}. This is particularly true for large geological domains which span several kilometres. It is possible to overcome these challenges by a family of techniques called {\it upscaling}. Upscaling techniques solve both uncertainty and scaling problems by replacing discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}.

Upscaling techniques which best preserve statistical properties and flow dynamics behaviour of the high resolution domain give the best representation of heterogeneous properties. There are several upscaling techniques which \red{fall} under different categories \cite{Hasting_2001, Renard_1997, Szymkiewicz_2013}. Traditional upscaling techniques are deterministic in nature, however over the last decade, techniques classed as stochastic have received great attention from academic and industrial porous media communities worldwide \cite{Guilleminot_2012, Ravalec-Dupin_2010, Verwoerd_2009}. One of the reasons for this is because stochastic upscaling techniques are robust enough to handle multiphase flow in porous media and they are able to address scaling and uncertainity problems.

{\it Upscaling} is a term commonly used in a few industries, and it is often referred throughout the applied mathematics and engineering communities as a form of model order reduction (MOR), or the output (\ie Reduced Order Model, ROM). Depending on the context in which MOR is being used, it can simply be referred to as dimensionality reduction. The definition of MOR also depends on the context it is being used \cite{Schilders2008}, but it suffices to state that MOR captures the essential features of a system, structure or parameter. This implies that an original system, structure or parameter exists with more detailed information. Fig.~\ref{fig:IllustrationMOR} presents a graphical intepretation of the MOR concept. It shows that even with the reduced number of facets, the graphic on the right of this illustration is recognized as that of a rabbit.
